---
title: An introduction to creating R packages
knit: (function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
    encoding=encoding,
    output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: Tom Wilson thomas.wilson@gov.scot
date: 5 October 2022
output: 
  ioslides_presentation:
    widescreen: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(devtools)
```

## What is an R Package?

- A package is what we install from [CRAN](https://cran.r-project.org/) (or an internal server) with `install.packages()` and load with `library()`.
- Bundles R code (functions, methods, classes) so usable by others.
- A package can include documentation, sample data, test scripts.
- It is surprisingly easy to create an R package and highly recommended!


## Why create an R Package?

- Sharing: Make processes you have developed easily accessible to others by installing the package, not copying code or script files.
- Reusability: Packages will usually bundle functionality that can be used across many projects.
- Reproducibility: Custom R packages may be used in, or developed to deliver a [Reproducible Analytical Pipeline](https://analysisfunction.civilservice.gov.uk/support/reproducible-analytical-pipelines/).
- Clarity: A known system for structuring an R code repository, documenting R processes, testing R code.
- Knowledge: Developing an R package gives a better understanding of how R works.

## An introduction to creating R Packages
### Aim to cover in this talk
x. Why create a package?
x. Useful reference materials and useful R libraries.
x. Basic run through of creating a package in R Studio.
x. Documenting functions within a package using `roxygen2`.
x. Checking a package for errors.
x. Building / compiling the package.

## An introduction to creating R Packages
### This introductory talk won't cover
x. Use of Git and GitHub.  
x. Formal code testing.
x. Creating vignettes.
x. Releasing a package on CRAN.
x. Rewriting R code in C++ and using Rcpp for performance.

## Reference materials

- [Creating R Packages](https://r-pkgs.org/) by Hadley Wickham and Jenny Bryan. Easy to follow introductory guide, based around [devtools](https://cran.r-project.org/web/packages/devtools/index.html) and related libraries.
- [Writing R Extensions](https://cran.r-project.org/doc/manuals/R-exts.html). Definitive guide on CRAN for reference.

## R Packages that are useful for creating a package
- [devtools](https://cran.r-project.org/web/packages/devtools/index.html) a bit like a 'tidyverse' for package development.
- [usethis](https://cran.r-project.org/web/packages/usethis/index.html) automating package setup and configuration tasks.
- [roxygen2](https://cran.r-project.org/web/packages/roxygen2/index.html) documentation and management of package metadata files.
- [testthat](https://cran.r-project.org/web/packages/testthat/index.html) formal unit testing of packages.


## Creating a package
```{r create-package, eval=FALSE}
usethis::create_package("rpackagetest")
```


The package name must:

- Be at least two characters
- Letters, numbers and . only (no other chracters including _ -)
- Start with a letter
- Can't end with a .
```{r check-available, eval=FALSE}
# Might use available package to check package name
available::available("mypackagename")
```

## Structure of an R package
- DESCRIPTION file provides important information about the package.
The title, version, authors, description can be edited.
Imported packages will later be added to this using `usethis::use_package("package_name")`.

- NAMESPACE contains the functions that will be available to users of the package.
- `R\` folder contains the code that will be created.

## Make package a git repository
```{r use-git, eval=FALSE}
usethis::use_git()
```

Does the same as `git init` but adds certain files to `.gitignore`.  
If want to put package on GitHub:  
- Easiest might be to create GitHub repo first of name of package.  
- Then can set the remote:    
`git remote add origin <github repo ssh link>`  
Alternatively, the `usethis` package does have a `use_github` function might explore.  
(Beware of main vs master if using older usethis package version).

## Create first function
```{r first-function, eval=FALSE}
usethis::use_r(clean_column_names.R)
```

- This simply creates a script to create a function under `R/`.  
- Often see each main function made available to uses of a package will be in its own R script under `R/`.


## What should go in the R/<script_name>.R files
- Note that only function definitions should be in these files.
- Do not use `library(package_name)` to load other libraries.
- Do not use `source("my_other_script.R")`.
- It is fine to have more than one function in one .R file, generally one main function and its helper function.
- See chapter 7 of [Building R Packages](https://r-pkgs.org/Code.html) for a discussion on these things.

## Create a documentation for the function.
In RStudio with cursor inside the function go to menu and select Code > Insert roxygen skeleton.  
Will see the roxygen comment using the `#'` combination with distinct sections defined by `@`. 
```{r roxygen, eval=FALSE}
#' Title
#'
#' @param df 
#'
#' @return
#' @export
#'
#' @examples
std_scale_df <- function(df) {
  df %>% dplyr::mutate_if(is.numeric, std_scale)
}
```

## Document a function - roxygen2
- The roxygen function comment generates the help for the function.
- Need to complete the template generated by Code > Insert roxygen skeleton:
  - A title describing the function in one line at the top.
  - If leave a blank line after the title sentence can then add more descriptive text until the first @.
  - A description of each input argument next to `@param`.
  - What it returns next to `@returns`.
  - An example of using it below the `@examples`.
- The `@export` means the function will be available to users once package loaded (adds to NAMESPACE). Don't write anything next to this.


## Importing other packages used in functions
- In most cases best to specify packages as `imports` not `depends`.
- If use `imports` then package is used within custom function like `dplyr::mutate`.
- If use `depends` then specified package is attached everytime.
- Helper function that adds package to `DESCRIPTION` > `imports`
```r
usethis::use_package("dplyr")
```

## Generate the documentation from roxygen comments
After populating the roxygen2 comments for exported functions, can generate the documentation.
```{r document-function, eval=FALSE}
devtools::document()
```

- Creates a `man/<function_name>.rd` documentation file
- Uses the `@export` tag to add functions that will be available to users in the NAMESPACE file.

## Load the package to test the functionality
```{r load-package, eval=FALSE}
devtools::load_all()
```
- This is simulating installing the package and loading it, so useful for checking while developing.
- Note the functions in the package will be available, but not seen in the global environment, just like when load with `library()`.

## Suggestions
- `devtools::check()` is a good way to ensure everything is in order.
- If need to use `%>%` or `:=` specify in roxygen function comment with `@importFrom`
```{r pipe, eval=FALSE}
#' @importFrom rlang :=
#' @importFrom magrittr %>%
```
- `usethis::use_pipe()` is an alternative way, setting a package so can use magrittr pipe `%>%` in every function.   
(definitely should not have `library(magrittr)` in package R files). 

## Distributing - building a package
- Use `devtools::build()`.
- If want a binary use `devtools::build(binary=TRUE)`.
- People can also install package from GitHub `devtools::install_github()`.
- The `.Rbuildignore` specifies files that will not be included in package tarball or binary. 
- Note that might wish to increment the version before building. This is in DESCRIPTION.
- However, there is a `usethis` function to do this:
```{r use_version, eval=FALSE}
usethis::use_version() #prompts with options of how to increment the version number.
```
