---
title: "Creating R Packages"
output: 
  ioslides_presentation:
    widescreen: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(devtools)
```
## An introduction to creating R Packages
### Aim to cover
x. Why create a package?
x. Useful reference materials and useful R libraries.
x. Basic run through of creating a package in R Studio.
x. Documenting functions within a package.
x. Checking a package for errors.
x. Building / compiling the package.

## An introduction to creating R Packages
### This talk won't cover
x. Use of Git and GitHub.  
x. Formal code testing.
x. Creating viginettes.
x. Using C and C++, Rccp for peformance.
x. Releasing a package on CRAN.

## What is an R Package?

- A package is what we install from [CRAN](https://cran.r-project.org/) (or an internal server) with `install.packages()` and load with `library()`.
- Bundles R code (functions) so easy to share with others.
- A package can include documentation, sample data, test scripts.
- It is surprisingly easy to create an R package and highly recommended!


## Why create an R Package?

- Sharing R code to be used by others:
  - Clear, known way for others to access your R code and start using it.
  - Best way to ensure consistent use of R functionality you've developed.
- Reproducibility and reusability from packaging code, so can be part of a [Reproducible Analytical Pipeline](https://analysisfunction.civilservice.gov.uk/support/reproducible-analytical-pipelines/).
- A clear workflow for structuring code repository, documenting processes, testing them.
- Creating an R package gives a better understanding of how R works.

## Reference materials

- [Creating R Packages](https://r-pkgs.org/) by Hadley Wickham and Jenny Bryan. Easy to follow introductory guide, based around [devtools](https://cran.r-project.org/web/packages/devtools/index.html) and related libraries.
- [Writing R Extensions](https://cran.r-project.org/doc/manuals/R-exts.html). Definitive guide on CRAN for reference.

## Packages that are useful for creating a package!
- [devtools](https://cran.r-project.org/web/packages/devtools/index.html) a bit like a 'tidyverse' for package development.
- [usethis](https://cran.r-project.org/web/packages/usethis/index.html) automating package setup and configuration tasks.
- [roxygen2](https://cran.r-project.org/web/packages/roxygen2/index.html) documentation and management of package metadata files.
- [testthat](https://cran.r-project.org/web/packages/testthat/index.html) formal unit testing of packages.


## Creating a package
```{r create-package, eval=FALSE}
usethis::create_package("rpackagetest")
```


The package name must:

- Be at least two characters
- Letters, numbers and . only (no other chracters including _ -)
- Start with a letter
- Can't end with a .
```{r check-available, eval=FALSE}
# Might use available package to check package name
available::available("mypackagename")
```
## Make package a git repository
```{r use-git, eval=FALSE}
usethis::use_git()
```
Does the same as `git init` but adds certain files to `.gitignore`.
If want to put package on GitHub:
- Easiest might be to create GitHub repo first of name of package.
- Then can set the remote:  
`git remote add orign <ssh link>` 
Alternatively, the `usethis` package does have a `use_github` function might explore.
(Beware of main vs master if using older usethis package version)

## Structure of a R package
- DESCRIPTION file provides important information about the package.
The title, version, authors, description can be edited.
Imported packages will later be added to this using `usethis::use_package("package_name")`

- NAMESPACE contains the functions that will be available to users of the package.
- `R\` folder contains the code that will be created.

## Create first function
```{r first-function, eval=FALSE}
usethis::use_r(clean_column_names.R)
```
This simply creates a script to create a function under `R/`.  

We want to create a documentation for the function.
In RStudio with cursor inside the function go to menu and select Code > Insert roxygen skeleton.

## What should go in the R/<script_name>.R files
- Note that only function definitions should be in these files.
- Do not use `library(package_name)` to load other libraries.
- Do not use `source("my_other_script.R")`.
- It is fine to have more than one function in one .R file, generally one main function and its helper function

## Importing other packages used in functions
- In most cases best to specify packages as `imports` not `depends`.
- If use `imports` then package is used within custom function like `dplyr::mutate`.
- If use `depends` then specified package is attached everytime.
- Helper function that adds package to `DESCRIPTION` > `imports`
```r
usethis::use_package("dplyr")
```

## Document a function - roxygen2
- In particular the main function should have roxygen2 comment.
- In RStudio, while cursor inside function, go to Code > Insert Roxygen Skelton
- This is the basis for the help for the function.
- Need to add:
  - Description of function at the top.
  - A description of each input argument.
  - An example of using.
- The `@export` means the function will be available to users once package loaded (adds to NAMESPACE).

## Generate the documentation
After populating the roxygen2 comments for exported functions, can generate the documentation.
```{r document-function, eval=FALSE}
document()
```

- Creates a `man/<function_name>.rd` documentation file
- Uses the `@export` tag to add functions that will be available to users in the NAMESPACE file.

## Load the package to test the functionality
```{r load-package, eval=FALSE}
usethis::load_all()
```
- This is simulating installing the package and loading it, so useful for checking while developing.
- Note the functions in the package will be available, but not seen in the global environment, just like when load with `library()`.

## Suggestions
- `check()` is a good way to ensure everything is in order.
- If need to use things like %>% and := can specify in function comment with `@importFrom`
#' @importFrom rlang :=
#' @importFrom magrittr %>%
- `use_pipe()` is a way to set a package so can use magrittr pipe `%>%` in function 
(definitely should not have `library(magrittr)` in package function.)

## Distributing - building a package
- Use the `build()`
- If want a binary use `build(binary=TRUE)`
- The `.Rbuildignore` specifies files that will not be included in package tarball or binary. 

