---
title: "Creating R Packages"
output: 
  ioslides_presentation:
    widescreen: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(devtools)
```

## What is an R Package?

- A package is what we install from [CRAN](https://cran.r-project.org/) (or an internal server) with `install.packages()` and load with `library()`.
- Bundles R code (functions) so easy to share with others.
- A package can include documentation, sample data, test scripts.
- It is surprisingly easy to create an R package and highly recommended!


## Why create an R Package?

- Sharing R code to be used by others:
  - Clear, known way for others to access your R code and start using it.
  - Best way to ensure consistent use of R functionality you've developed.
- Reproducibility from packaging code makes it an important part of a [Reproducible Analytical Pipeline](https://analysisfunction.civilservice.gov.uk/support/reproducible-analytical-pipelines/).
- A clear workflow for structuring code repository, documenting processes, testing them.
- Creating an R package gives a better understanding of how R works.

## Reference materials

- [Creating R Packages](https://r-pkgs.org/) by Hadley Wickham and Jenny Bryan. Easy to follow introductory guide, mostly using [devtools](https://cran.r-project.org/web/packages/devtools/index.html)
- [Writing R Extensions](https://cran.r-project.org/doc/manuals/R-exts.html). Definitive guide on CRAN for reference.

## Packages used when creating a package!
- [devtools](https://cran.r-project.org/web/packages/devtools/index.html) has many functions to make all stages of package development easier and this is what will be used in this demo.
- [usethis](https://cran.r-project.org/web/packages/usethis/index.html) is the package with much of this functionality imported by devtools.
- [roxygen2](https://cran.r-project.org/web/packages/roxygen2/index.html) documentation and metadata files when creating a package
- [testthat](https://cran.r-project.org/web/packages/testthat/index.html) another Hadley Wickham package used for formal testing of packages.


## Creating a package
```{r create-package, eval=FALSE}
library(devtools) # this loads usethis which is source of many of the functions shown
create_package("rpackagetest")
```


The package name must:

- Be at least two characters
- Letters, numbers and . only (no other chracters including _ -)
- Start with a letter
- Can't end with a .

## Make package a git repository
```{r use-git, eval=FALSE}
use_git()
```
(see later slide on sharing regarding putting package on GitHub)

## Structure of a R package
- DESCRIPTION file provides important information about the package.
The title, version, authors, description can be edited.
Imported packages will later be added to this using `usethis::use_package("package_name")`

- NAMESPACE contains the functions that will be available to users of the package. 
Recomm

## Create first function
```{r first-function, eval=FALSE}
use_r(clean_column_names.R)
```
This simply creates a script to create a function under `R/`.  

We want to create a documentation for the function.
In RStudio with cursor inside the function go to menu and select Code > Insert roxygen skeleton.

## Generate the documentation
```{r document-function, eval=FALSE}
document()
